---
title: "p8105_hw1_daq2107.Rmd"
author: "Diego A. Quintana Licona (daq2107)"
date: "2026-09-23"
output: github_document
---

## Loading necessary libraries for this assignment:

##### what does the descriptions within {} do???

##### ask about RData file...

```{r libraries}
library(tidyverse)
library(skimr)
```

## Problem 1

Loading 'penguins' data set:

```{r loading_data}
data("penguins", package = "palmerpenguins")
```

After the data has been loaded, we can explore the 'penguins' data set:

```{r exploring_penguins}
penguins #General observation of the data frame

nrow(penguins) #Obtaining number of rows
ncol(penguins) #Obtaining number of columns

skim(penguins$flipper_length_mm) #Using the 'skimr' package to obtain mean flipper length

```

#### what do we say about 'important values of variables'?
#### okay to use 'skmr' package or do we need to just use mean?

**Description:** We see that the 'penguins' data set is a tibble containing 344 observations and 8 variables, as confirmed by the 'nrow' and 'ncol' functions (344 rows and 8 columns). Each observation corresponds to a penguin, whereas the variables describe their location ('island' variable), their species, sex, year of  physical characteristics of each penguin, including their bills' length and depth (mm), their flippers' length (mm), and overall body mass (g).

Additionally, we see that the mean flipper length for all observations in the data set is 200.9152 mm.

We can also create a scatter plot of flipper length (y) vs. bill length (x):

```{r scatter_plot_penguins}
#Using ggplot (part of 'tidyverse' package):

ggplot(penguins, aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +  #Coloring per 'species' variable
    geom_point() +
  
#Adding labels to my scatter plot (I know this code per previous experience with ggplot)
  
  labs(title = "Penguin Flipper Length vs. Bill Length",
       x = "Bill Length (mm)",
       y = "Flipper Length (mm)") +
  
  theme(plot.title = element_text(hjust = 0.5)) #Centering my title

#Saving my plot using 'ggsave':

ggsave(filename = "DSI_HW1_Problem1.jpg")

```

##### jpg or png???
#### do we need to address the missing values???
#### okay extra ggplot things considering my previous class????
#### email reminding him older R version. to Gustavo???

## Problem 2

We can create a data frame using the 'tibble' function (part of the 'tidyverse' package):

####### do we need to take 'seed'? to make sure means remain consistent?


```{r creating_df}
new_df = tibble(
  vec_sample = rnorm(10, mean = 1), #Random sample of 10 and mean equal to 1
  vec_log = vec_sample > 0, #Logical vector checking if the sample's elements are above zero
  vec_char = c("This", "is", "my", "ten", "character", "vector", "with", "great", "joy", "woohoo!"), #Character vector
  vec_factor = factor(c("Mexican", "Japanese", "Japanese", "Mexican", "Lebanese", "Lebanese", "Mexican", "Mexican", "Lebanese", "Japanese")) #Factor vector with three nationality levels
)
```

After we have created our desired data frame, we can attempt to take the mean of each newly-created variable:

```{r taking_means}

mean(pull(new_df, vec_sample))
mean(pull(new_df, vec_log))
mean(pull(new_df, vec_char))
mean(pull(new_df, vec_factor))

```

We can see from the output above that we are able to take the mean for our random sample (mean = 0.7015389) and logical vectors (mean = 0.7). However, as expected, we cannot take the means of our character and factor vectors since they are not numerical.

We can attempt to use the 'as.numeric' function to convert the variables into numerical variables and then take the mean:

```{r convert_numeric}

as.numeric(new_df$vec_log)
as.numeric(new_df$vec_char)
as.numeric(new_df$vec_factor)

mean(as.numeric(pull(new_df, vec_log)))
mean(as.numeric(pull(new_df, vec_char)))
mean(as.numeric(pull(new_df, vec_factor)))

```

####### confirm in office hours
####### says to use chunk but not output???

As we can see, the 'as.numeric' function was able to convert the logical vector elements into 0 or 1 (values corresponding to 'TRUE' or 'FALSE') and the factor vector elements to 1, 2, or 3 (values corresponding to each of the three levels). The means for the logical and factor vectors, after converting to numeric, we see they have means of 0.7 and 2.1, respectively. However, we see that the character vector values still remain as 'NA', and so does the mean value. This makes sense since there is no standardized logic or order for character elements to convert them to numerical elements.

